\subsection{Isolation Specifications}
\label{sec:ansi-isolation}

\begin{figure*}[t]
\begin{smathpar}
\begin{array}{lcl}
(\A,\visZ) \Vdash \eta \in T_i & \Leftrightarrow & \eta \in \A \conj \txn(\eta) = T_i\\
(\A,\visZ) \Vdash \eta_1 \visar \eta_2 & \Leftrightarrow & \{\eta_1,\eta_2\}
        \subseteq \A \conj (\eta_1,\eta_2) \in \visZ\\
(\A,\visZ) \Vdash \eta_1 \soar \eta_2 & \Leftrightarrow & \{\eta_1,\eta_2\}
        \subseteq \A \conj \txn(\eta_1)=\txn(\eta_2) \conj \id(\eta_1)
        < \id(\eta_2)\\
\E \Vdash S \subseteq T_i & \Leftrightarrow & \forall \eta.~ \eta
        \in S \Rightarrow \underE{\eta \in T_i} \\
\E \Vdash T_i \visar \eta & \Leftrightarrow &\forall\eta_1
        .\,(\E \Vdash \eta_1 \in T_i) \Rightarrow \E \Vdash \eta_1 \visar \eta \\
\underE{T_i \visar T_j} & \Leftrightarrow &  \forall\eta_1,\eta_2.\,
    %\sameobj{\eta_1}{\eta_2}  \Rightarrow 
    \underE{\eta_1\in T_i} \conj \underE{\eta_2 \in T_j} \Rightarrow 
    \underE{\eta_1 \visar \eta_2} \\
\underE{T_i \invisar T_j} & \Leftrightarrow &  \forall\eta_1,\eta_2.\,
        %\sameobj{\eta_1}{\eta_2}\Rightarrow 
        \underE{\eta_1\in T_i} \conj \underE{\eta_2 \in T_j} \Rightarrow 
        \neg (\underE{\eta_1 \visar \eta_2})\\
\underE{T_i \wrstoar X} & \Leftrightarrow & \exists\eta.~
        \underE{\eta \in T_i} \conj \kind(\eta) = \C{WR}(X)\\
\underE{\C{RMWVis}(T_j)} & \Leftrightarrow & \forall\eta_1,\eta_2.\,
       \underE{\{\eta_1,\eta_2\} \subseteq T_j} \conj
       \underE{\eta_1 \soar \eta_2} \Rightarrow \underE{\eta_1 \visar
       \eta_2}\\
\underE{\C{MonotonicVis}(T_j)} & \Leftrightarrow & 
       \underE{\C{RMWVis}(T_j)} \conj 
       \forall\eta,\eta_1,\eta_2.\, \underE{\{\eta_1,\eta_2\} \in T_j} 
          \conj \\
  &   & \hspace*{1.2in}\underE{\eta \visar \eta_1} \conj
        \underE{\eta_1 \soar \eta_2} \Rightarrow \underE{\eta \visar
        \eta_2} \\
%  \C{CausalVis}(T_i) & \Leftrightarrow & 
%         \C{MonotonicVis}(T_i) \conj \C{RMWVis}(T_i)\\
\underE{\C{AtomicVis}(T_j)} & \Leftrightarrow & 
       \forall\eta_1,\eta_2.\, \neg(\underE{\eta_1 \in T_j}) \conj
       \underE{\eta_2 \in T_j} \conj
       \underE{\eta_1 \visar \eta_2} \Rightarrow \underE{\txn(\eta_1)
       \visar \eta_2}\\
\underE{\C{CommitVis}(T_j)} & \Leftrightarrow & 
       \forall\eta_1,\eta_2.~ \neg(\underE{\eta_1 \in T_j}) \conj 
          \underE{\eta_2 \in T_j} \conj
       \underE{\eta_1 \visar \eta_2} \Rightarrow\\
  &   & \hspace*{1.2in}\exists\eta.\, \underE{\eta \in \txn(\eta_1)} 
        \conj \kind(\eta) = \C{COMMIT} 
        \conj \underE{\eta \visar \eta_2}\\
% \underE{\C{TransVis}(T_j)} & \Leftrightarrow &  \forall
%        \eta_1,\eta_2,\eta_3.\, \underE{\eta_3 \in T_j} \conj
%        \underE{\eta_1 \visar \eta_2} \conj
%        \underE{\eta_2 \visar \eta_3} \Rightarrow \underE{\eta_1 \visar
%        \eta_3} \\
\underE{\C{SnapshotVis}(T_i,T_j)} & \Leftrightarrow &  \underE{T_i
       \visar T_j} \disj \underE{T_i \invisar T_j}\\
\underE{\C{OneWaySER}(T_i,T_j)} & \Leftrightarrow &  \underE{T_i
       \visar T_j} \disj (\underE{T_i \invisar T_j} \conj \\
  &   &\hspace*{1.2in}\exists \eta.~\underE{\eta\in T_i} \conj \kind(\eta) =
       \C{COMMIT} \Rightarrow \underE{T_j \visar \eta})\\
\underE{\C{RC}(T_j)} & \Leftrightarrow & \underE{\C{AtomicVis}(T_j)} 
        \conj \underE{\C{CommitVis}(T_j)}\\
\underE{\C{MAV}(T_j)} & \Leftrightarrow & \underE{\C{RC}(T_j)} \conj
      \underE{\C{MonotonicVis}(T_j)} \\
\underE{\C{RR}(T_j)} & \Leftrightarrow & \underE{\C{MAV}(T_j)}
       \conj \forall T_i.\,T_i \neq T_j \Rightarrow 
        \underE{\C{SnapshotVis}(T_i,T_j)} \\
% &   & \hspace*{2in}\conj  \C{SnapshotVis}(T_i,T_j)\\
\underE{\C{SI}(T_j)} & \Leftrightarrow &  \underE{\C{RR}(T_j)}
       \conj \forall T_i.\,(T_i \neq T_j \conj \exists X.\, \underE{T_i \wrstoar X} \conj 
        \underE{T_j \wrstoar X})\\
%      (\exists X.\, T_i \wrstoar X \conj T_j \wrstoar X)
%      \Rightarrow  \C{TotalVis}(T_i,T_j)\\
  &   & \hspace*{2in} \Rightarrow \underE{\C{OneWaySER}(T_i,T_j)}\\
\underE{\C{SER}(T_j)} & \Leftrightarrow & \underE{\C{MAV}(T_j)}
       \conj \forall T_i.\,T_i \neq T_j 
       \Rightarrow \underE{\C{OneWaySER}(T_i,T_j)}\\
% &   & \hspace*{2in}\conj  \C{TotalVis}(T_i,T_j)\\
\end{array}
\end{smathpar}

\caption{Standard isolation guarantees expressed as trace
well-formedness constraints}
\label{fig:ansi-isolation}
\end{figure*}

Fig.~\ref{fig:ansi-isolation} shows the specification of standard
isolation guarantees expressed as constraints over trace
well-formedness. For brevity and convenience, we adopt some notation.
An execution trace is destructed into $\A$ and $\visZ$ whenever
individual components of the pair are needed.  Otherwise, it is written
as $\E$. When $\psi$ is a proposition, we write $\underE{\psi}$ to denote
an interpretation of $\psi$ in the context of the trace $\E$. Such
interpretations are defined on a case-by-case basis in
Fig.~\ref{fig:ansi-isolation}. In the following, we give informal
explanations for each definition.

In the context of a trace $(\A,\visZ)$, an effect $\eta$ is said to
belong to a transaction $T_i$ if $\eta$ belongs to the effect set $A$
and its transaction identifier is $T_i$. The containment relation is
trivially lifted to the set of effects to define $\underE{S \subseteq
T_i}$.  The interpretation of $\eta_1 \visar \eta_2$ and $\eta_1
\soar \eta_2$ are straightforward and explained in \S\ref{sec:syntax}.
A transaction $T_i$ is said to be visible to an effect $\eta$ if every
effect $\eta_1$ of $T_i$ recorded by the trace is visible to $\eta$.
$T_i$ may be visible to $\eta$ but may not be visible to every other
effect in the transaction. For a transaction $T_i$ to be considered to
be visible to a transaction $T_j$ in the context of a trace $\E$
(written $\underE{T_i \visar T_j}$), every effect ($\eta_1$) of $T_i$
present in $\E$ must be visible to every effect ($\eta_2$) of $T_j$ in
$\E$. Conversely, if none of the effects of $T_i$ present in $\E$ are
visible to any effect of $T_j$, then $T_i$ is considered invisible to
$T_j$ under $\E$ (written $\underE{T_i \invisar T_j}$). Transaction
$T_i$ is said to have written to a variable $X$ under $\E$ (i.e.,
$\underE{T_i \wrstoar X}$) if there exists a write-to-$X$ effect from
$T_i$ in $\E$.

Various isolation guarantees are defined as propositions indexed by
a transaction identifier $T_j$. Transaction $T_j$ is said to have
experienced \emph{read-my-writes} visibility under $\E$ if every
effect ($\eta_1$) of $T_j$ is visible to every subsequent effect
($\eta_2$) of the same transaction in $\E$. This lets $T_j$ to never
lose its own updates. Monotonic visibility adds one more constraint to
read-my-writes; besides requiring $\eta_1$ to be visible to $\eta_2$,
it also requires every effect ($\eta$) visible to $\eta_1$ in $\E$ to
be visible to $\eta_2$ as well. Thus $T_j$ witnesses monotonically
increasing state as time progresses. Atomic visibility allows an
effect $\eta_2$ of $T_j$ to witness an effect $\eta_1$ of $T_i$ only
if all effects of $T_i$ in $\E$ are also visible to $\eta_2$. Atomic
visibility thus prevents a transaction from being partially visible.
However, atomic visiblity does not prevent an uncommitted transaction
from being visible. To confine visibility to only committed
transactions, we need an additional constraint that requires an effect
($\eta_2$) of $T_j$ to also witness the commit effect of $T_i$
whenever it witnesses some effect ($\eta_1$) of $T_i$. This additional
constraint is captured by the $\mathtt{CommitVis}$ specification. A
transaction $T_i$ is said to be snapshot-visible to a transaction
$T_j$ if either it is visible to $T_j$, or it is invisible; it is
forbidden for only a suffix (more generally, a subset) of $T_j$ to
witness $T_i$. Observe that snapshot visibility permits $T_i$ and $T_j$ to
execute and commit while being  oblivious of each other. One-way
serializability proscribes this possiblity. $T_i$
is said to be one-way-serializable w.r.t $T_j$ if either it is visible
to $T_j$ or $T_j$ is visible to the commit effect of $T_i$.  Note that
both $\mathtt{SnapshotVis}$ and $\mathtt{OneWaySER}$ are asymmetric
definitions that guarantee complete visiblity or invisibility of $T_i$
to $T_j$, but not the converse. While $\mathtt{SnapshotVis}$ provides
no guarantees to $T_i$, $\mathtt{OneWaySER}$ guarantees that at least
the commit effect of $T_i$ witnesses $T_j$.

The ANSI SQL 92 standard requires \iso{Read Committted} isolation to avoid
the dirty reads phenomenon, which is achieved by enforcing
$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ guarantees. The {\sc rc}
specification is therefore a combination of these two guarantees. The
specification also agrees with the description and
implementation~\cite{bailishat,pldi15} of {\sc rc} for highly available
replicated stores. On relational databases, however, {\sc rc} has also come
to be associated with the $\mathtt{MonotonicVis}$ guarantee.
Nonetheless, $\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ are
sufficient to reason about {\sc rc} isolation on relational stores too. As
we demonstrate in \S\ref{sec:store-consistency}, the combination of
these guarantees with the {\sc sc} property of relational stores
automatically leads to the monotonicity guarantee, which probably
explains why {\sc rc} comes with $\mathtt{MonotonicVis}$ on such stores. On
weakly consistent stores however, $\mathtt{AtomicVis}$ and
$\mathtt{CommitVis}$ do not imply $\mathtt{MonotonicVis}$. A stronger
isolation level called \iso{Monotonic Atomic View} ({\sc mav} of
Fig.~\ref{fig:ansi-isolation})~\cite{bailishat,pldi15} was proposed to
explicitly extend {\sc rc} with monotonicity on such stores. \iso{Repeatable
Read} spec ({\sc rr}) extends {\sc mav} with the snapshot visibility guarantee
as described before. \iso{Snapshot Isolation} spec ({\sc si}) extends
{\sc rr} with a one-way serializability guarantee w.r.t the transactions
that perform conflicting writes (i.e., writes to the same shared
variable). Finally, \iso{Serializable} isolation extends one-way
$\mathtt{SnapshotVis}$ to all transactions, including those that do
not conflict\footnote{Fig.~\ref{fig:ansi-isolation} presents slightly
weaker versions of {\sc si} and {\sc ser} specs in the interest of
clarity.}. 

In contrast to recent proposals (\emph{e.g.},
~\cite{gotsmanconcur15}), our specifications for {\sc si} and {\sc
  ser} do not necessarily impose a total order among transactions
(conflicting or otherwise). In reality, a total order under {\sc ser}
(resp. {\sc si}) is guaranteed only if the store executes all
transactions under {\sc ser} (resp. {\sc SI}) isolation. Our
specifications admit this possibility, and derive a total order under
the assumption of homogenity. However, databases almost always allow
isolation levels to be configured on a per-transaction basis, allowing
transactions at various isolation levels to coexist.  Specifications
that assume homogenity are incorrect under this setting.

% Having specified isolation levels as trace well-formedness
% constraints, we can now construct trace invariants ($\I$) for
% \txnimp programs by composing isolation specifications for various
% transactions. For instance, the following trace invariant enforces
% {\sc si} for both transactions of the program in
% Fig.~\ref{fig:motiv-eg-1}, allowing it to satisfy its postcondition:
% \begin{smathpar}
% \I \;=\; \lambda\E.~ \underE{\C{SI(Wd1)}} \conj \underE{\C{SI(Wd2)}}
% \end{smathpar}
% In contrast, the following trace invariant enforces {\sc rc} for one and {\sc si}
% for another, leading to a possible violation of the postcondition:
% \begin{smathpar}
% \I \;=\; \lambda\E.~ \underE{\C{RC(Wd1)}} \conj \underE{\C{SI(Wd2)}}
% \end{smathpar}
