\section{\txnimp}
\label{sec:syntax}

\subsection{\txnimp: Syntax and Semantics}
\label{sec:opsem}

\input{txnimp}

Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
\txnimp, a core language that we will use to formalize the intuitions
presented in the previous section. Natural numbers, (shared) variables
and arithmetic expressions constitute the syntactic class of
expressions ($e$).  Commands ($c$) include $\cskip$, assignment
statements, transaction (\C{txn}) lexical blocks, and their sequential
and parallel composition. We let $T_i$ for $i \in \mathbb{N}$ range
over transaction identifiers. When it is evident we are referring to a
transaction, we use the number $i$ instead of $T_i$ for identification
(\eg, in $\C{txn}\langle i \rangle$). For notational convenience, we
let $t$ range over both expressions and commands.

We define a small-step operational semantics for this language in
terms of an abstract machine that generates an execution trace
($\E$). The first component of the trace is a set ($\A$) of
\emph{effects}, where an effect ($\eta$) documents a read (\C{RD(X)}),
a write (\C{WR(X)}), or a transaction commit (\C{COMMIT}) operation
performed during the execution. Any value associated with the
operation (\eg, value read or value written) is also documented, and
can be accessed via $\rval$. Every effect has a unique identifier
accessible via $\id$, and a transction identifier accessible via
$\txn$.  The latter identifies the transaction that generated the
effect. In every step of the evaluation, the machine reduces a \txnimp
term by executing a read, write or commit operation, generating an
effect, and extending the trace. Since effects include transaction
identifiers, the semantics distinguishes between terms ($t$) of
different transactions. For example, $\txnbox{t}_i$ denotes a term $t$
inside a transaction $T_i$.  Evaluation contexts are also
appropriately marked. For example, $\ectx_i$ denotes the evaluation
context for a term inside $T_i$. The other component of an execution
trace is a visibility relation ($\visZ$) that establishes a visibility
property between effects among different transactions. The intent and
mechanics of $\visZ$ is described in the sequel.

Fundamental to our development is the notion of a trace invariant
($\I$). $\I$ is a function from traces ($\E$) to propositions
($\texttt{Prop}$) that define well-formedness constraints over traces.
The machine takes a step only if the resulting trace satisfies the
constraints imposed by $\I$. This behaviour is captured by the
auxiliary reduction rule \rulelabel{E-Aux} that factors out the trace
extension aspect of the evaluation by abstracting away the
operation-specific behaviour as a function that generates an
appropriate effect. We let $\mathcal{F}$ denote this function.
\rulelabel{E-Aux} uses $\mathcal{F}$ to generate a new effect and
extend the trace ($\E = (\A,\visZ)$) \emph{only if} the
well-formedness constraints imposed by $\I$ on $\E$ (i.e., $\I(\E')$)
are satisfied. Otherwise, it gets stuck. In an execution that runs to
completion, every small-step preserves the well-formedness of a trace,
thus ensuring the invariance of $\I$.  Note that the semantics makes
no assumptions about $\I$ other than its type. As such, it can be
instantiated with any trace-parametric proposition that expresses
constraints over the given trace. For instance, consider the
$\psi_{RC}$ specification from \S\ref{sec:motivation}, but with
bounded $T_1$ and $T_2$ instantiated with \C{Wd1} and \C{Wd2},
respectively. The instantiated specification is the following
proposition:

\begin{smathpar}
\begin{array}{l}
  \forall \eta_1,\eta_2,.\; \txn(\eta_1) = \C{Wd1}
  \conj \txn(\eta_2) = \C{Wd2} \\
  \hspace*{0.6in}\conj \C{Wd1} \neq \C{Wd2} \conj \eta_1 \hboar
  \eta_2 \Rightarrow \C{Wd1} \hboar \eta_2 \\
\end{array}
\end{smathpar}

\noindent It is easy to intepret the above proposition in the context of a trace
$\E$ that captures an execution of the program in
Fig.~\ref{fig:motiv-eg-1}. Such a trace-parametric proposition can be
used to instantiate the trace invariant $\I$ in Fig.~\ref{fig:txnimp}.
The resultant operational semantics describes an abstract machine that
gets stuck if an operation of \C{Wd2} is executed in a state that
incorporates some, but not all the effects (including the \C{COMMIT})
of \C{Wd1}. 
% While the machine takes a step only if the constraints are
% satisfied, it neither defines nor explicitly assumes an oracle to
% check satisfaction.

As described in \S\ref{sec:motivation}, the semantics of various
isolation levels can be captured as constraints over the
happens-before ($\hbZ$) relation. $\hbZ$ is however a derived relation
in our model, composed of more fundamental \emph{session order}
($\soZ$) and \emph{visibility} ($\visZ$) relations, which shall
henceforth replace $\hbZ$ in the discussion.  A session order relation
captures the sequential order of operations within a transaction. In
particular, it relates two effects, $\eta_1$ and $\eta_2$, such that
$\txn(\eta_1) = \txn(\eta_2)$ and $\id(\eta_1) < \id(\eta_2)$.  The
semantics assigns monotonically increasing identifiers to effects, as
defined by the $\id(\eta) > \maxId(\A)$ condition of
\rulelabel{E-Aux}.  Evaluation contexts ($\ectx_i$) for
transaction-bound terms are defined so as to enforce a deterministic
sequential order of execution within a transaction, leading to a
deterministic total order among effect ids, which defines the session
order relation. Visibility ($\visZ$) on the other hand relates effects
across concurrent transactions.  Intuitively, $\visZ$ relates $\eta_1$
to $\eta_2$ if and only if $\eta_1$ was \emph{visible} to the
operation that generated $\eta_2$ during its execution, thus effecting
its return value ($\rval(\eta_2)$). For example, a read operation over
\C{X} may pick the value ($\rval$) of the write effect with highest id
among the visible effects (this is made possible by appropriately
defining $\interp{\cdot}$ in \rulelabel{E-Var}, as we show later).
Thus, the value of a read depends on what write effects it can
witness. An operation can only witness the effects of already
concluded operations, which varies between executions due to the
non-deterministic order of evaluating the parallel composition of
transactions. A more notable source of non-determinism, however, is
the \rulelabel{E-Aux} rule, which allows the machine to expose an
arbitrary subset ($S$) of existing effects ($\A$) to the incoming
operation. In other words, the machine is not obligated to reveal the
effects of all previous operations to an incoming operation. This
relaxation allows the abstract machine to model the semantics of
weakly consistent data stores. For instance, operations issued to an
eventually consistent (EC) replicated store could be dispatched to
different replicas whose states may not be in any well-defined
relationship. By allowing operations to witness arbitrary subsets of
the global state, the semantics models the weak visibility properties
of such stores. Stronger visiblity properties needed for stronger
isolation can however be recovered by imposing well-formedness
constraints over $\visZ$ via the trace invariant ($\I$). Since the
abstract machine is obligated to satisfy $\I$ at every step of the
execution, operations are guaranteed to experience the level of
isolation specified by $\I$.  Thus, in executions that run to
completion, the abstract machine models a store that provides the
required levels of isolation.  Notably, the machine achieves this
without defining an operation semantics for isolation levels, solely
by relying on their declarative characterization as trace
well-formedness constraints.  \S\ref{sec:ansi-isolation} shows the
specification of various ANSI SQL isolation levels stated as trace
well-formedness constraints.

% However, a machine that lets operations witness arbitrary subset of
% the global state offers no isolation whatsoever. For example, it may
% allow a read operation to witness writes of an uncommitted
% transaction, violating RC isolation. Fortunately, our ability to
% express an isolation semantics as constraints over happens-before
% order through $\visZ$ and $\soZ$ relations, and the property of the
% abstract machine to be parametric over the trace invariant ($\I$),
% lets us solve this problem.
% In particular, we continue to define isolation semantics as
% constraints over $\visZ$ and $\soZ$, but confine their domain of
% interpretation to the given trace so that they now become trace
% well-formedness constraints. Well-formedness constraints can be
% combined into a trace invariant ($\I$). 


As described previously, \rulelabel{E-Aux} abstracts away the
operation-specific behaviour of a machine step as a function ($\F$)
that accepts a set ($S$) of effects chosen by the machine to make
visible to the operation, interprets the operation w.r.t $S$, and
returns an appropriate effect that encodes its return value. Rules
\rulelabel{E-Var}, \rulelabel{E-Asgn} and \rulelabel{E-Commit} define
such functions for read, write and commit operations, respectively.
The effect returned by the function in each case includes its
transaction id ($T_i$) along with an arbitrarily chosen effect id
($j$) that is later verified to be unique in \rulelabel{E-Aux}. The
$\rval$ for a write is the value being written, and for commit it is
$\bot$. In case of a read, the value read depends on the how the read
operation chooses to interpret the given set ($S$) of visible effects.
The interpretation may depend on the application semantics. For
example, a monotonically increasing counter application may choose to
let a write with the largest value determine the value of a
read. A more straightforward interpretation has already been described
before. To accommodate multiple interpretations, the semantics
is made parametric over an interpretation function ($\interp{\cdot}$)
that accepts a set of effects and a variable name, and returns the
value associated with the variable. A straightforward interpretation
function that chooses the last write (i.e., write with largest id) is
shown below:

\begin{smathpar}
\begin{array}{lcl}
  \isMax(S,\eta) & \Leftrightarrow &  \forall (\eta'\in S).  
  \kind(\eta') = \kind(\eta) \\
  & & \hspace*{0.4in}\Rightarrow \eta' = \eta \disj \id(\eta') < \id(\eta)\\

\interp{S}(X) & = & \C{if}\;(\exists (\eta \in S). \kind(\eta) = \C{WR}(X) 
  \wedge \isMax(S,\eta)) \\
  & & \C{then}\;\rval(\eta)\;\C{else}\;0\\
\end{array}
\end{smathpar}

\noindent Rules \rulelabel{E-Top-Ctx} and \rulelabel{E-Txn-Ctx} define
congruence properties for top-level terms and transaction-bound terms,
respectively. The rules and evaluation contexts ($\ectx$ and
$\ectx_i$) are defined such that only certain kinds of terms are
allowed at the top-level and inside a transaction. In particular, a
\txnimp program at the top-level can either be a transaction, or a
parallel composition of transactions. A command inside a \C{txn} block
can either be an assignment, or a sequential composition of
assignments. 

\input{ansi-isolation}
